---
title: "Data Analysis of EPA CO Daily Summary"
author: "Daniel Paliura"
date: "5/15/2021"
output: pdf_document
urlcolor: brown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

This document drawn up to research the data in EPA CO Daily Summary data set.
This research must set up the ground for mathematical model of forecasting
program. Here I have to find which factors affect on CO amounts measured.
For example, I expect that location of sites and probably measurements method
affect values measured. 

I will try to answer questions asked in *Exploratory analysis*, they are bold
font. And additionally to find some patterns in data.

The main goal is to determine which data I can use in forecasting and how
models for different factors could differs.

### Questions

1. How amount of measuring sites differs in different years?
2. Does number of monitors at same sites (unique POC count) changes in time?
3. Why ```poc``` values 6, 7, and 8 aren't present?
4. Why there are more values with ```poc``` value 9 than for values 4 and 5?
5. Whether event at some day is written into each observation at same day?
6. Whether data significantly differs by event type factor (regression analysis)?
7. Do events change values in perspective? 
8. Would exceptional event presence increases forecasts error compared to
    forecasting without such event?
9. Why amount of ```observation_percent``` value equals to 8 is greater than
    amounts of neighbor values 4, 13, 17, 21, 25?
10. Why feature ```arithmetic_mean``` contains negative values?
11. Are values of ```arithmetic_mean``` distributed (log)normally for separate 
    sites/countries/states?
12. Are negative values of ```first_max_value``` dependent of some factor?
    Measuring method for example
13. Whether all 1-hour methods has not available AQI?
14. Is there any significant differences in measurements distributions between
    different methods?
15. How do measured values differs in different states?
16. Does distributions significantly different by factor ```cbsa_name```?
17. Are measurements made with NDIR method significantly different by factor
    ```method_code``` inside groups NDIR and NDIR PHOTOMETRY? And hence can same
    methods with different codes be merged?
18. What the result of two-way AnoVa on factor ```polutant_standard```?

## Analysis

### Preparations

I use following R packages:

```{r source libraries}
library(dplyr, warn.conflicts = FALSE)
library(lubridate, warn.conflicts = FALSE)
```

And read data. I will use here data set connected with codes, so I will have
codes of states and counties and also codes. So I won't be forced to restore
relations to determine method or state or county.

```{r read data}
folder <- "../data/parted/by_codes/"

na.strs <- c('NA', '', '-')

obs <- read.csv(paste0(folder,"observations.csv"), na.strings=na.strs)[,-1]
sit <- read.csv(paste0(folder,'sites.csv'), na.strings=na.strs)[,-1]
met <- read.csv(paste0(folder,'methods.csv'), na.strings=na.strs)[,-1]

rm(folder, na.strs)
```

Now let's begin answering the questions.

### Question 1

*How amount of measuring sites differs in different years?*

```{r question 1 answer, echo=FALSE}
sites_amount_in_years <- obs %>%
    mutate(year = year(date_local)) %>%
    group_by(year) %>%
    summarise(n_sites=n_distinct(state_code, county_code, site_num, datum))

sites_amount_in_years

plot(sites_amount_in_years,
     ylim = c(0, max(sites_amount_in_years$n_sites)),
     type = "b", pch = 20,
     main = "Dependence of measuring sites amounts in different years")
```

**Number of measuring sites changing from year to year.**
**First years number of sites was increasing.**
**After 1996 number of sites begun decreasing.**
**It declined by third part after 2010 year.**

It would be interesting to see how much sites didn't stop working or
worked more than half of whole period.

### Question 2

*Does number of monitors at same sites (unique POC count) changes in time?*

To answer this question I have to group all observations by sites and dates and
count unique ```poc```.

```{r question 2 answer, echo=FALSE}
n_poc_per_date <- obs %>%
    group_by(state_code, county_code, site_num, datum, date_local) %>%
    summarise(n_poc = n_distinct(poc))

n_poc_summary <- n_poc_per_date %>%
    group_by(state_code, county_code, site_num, datum) %>%
    summarise(n_min_monitors = min(n_poc),
              n_max_monitors = max(n_poc))

cat("Preview of monitor amounts summary per sites.\n",
    "n_min_monitors is minimum number of monitors at the site over the entire period.\n",
    "n_max_monitors is maximum number of monitors at the site over the entire period.\n",
    sep="")
rbind(
    head(n_poc_summary, 3),
    tail(n_poc_summary, 3)
)

cat("Unique values of n_min_monitors:\n")
unique(n_poc_summary$n_min_monitors)
cat("Unique values of n_max_monitors:\n")
unique(n_poc_summary$n_max_monitors)

cat("Next table shows how many sites had had which minimum and maximum of monitors\n")
n_poc_summary %>% 
    group_by(n_min_monitors, n_max_monitors) %>%
    summarise(sites_amount = n())

```

So now we know that
**number of monitors is changeable, but it's a rare phenomenon**.
Only 14 of 1299 sites had had maximum 2 monitors through entire period.
And only 2 sites had had maximum 3 monitors measuring CO.
It's pretty small amount of monitors.

### Question 3

*Why POC values 6, 7, and 8 aren't present?*

```{r answer to question 3, echo=FALSE}
sites_with_many_monitors <- n_poc_summary %>%
    filter(n_max_monitors>1) %>%
    select(-n_min_monitors)

cond <- logical(nrow(obs))
for (i in 1:nrow(sites_with_many_monitors)){
    site <- sites_with_many_monitors[i,]
    
    cond <- cond | (obs$state_code  == site$state_code &
                    obs$county_code == site$county_code &
                    obs$site_num    == site$site_num &
                    obs$datum       == site$datum)
}
obs_with_many_monitors <- 
obs_with_singl_monitor <- obs[!cond,]

cat("Unique POC values for sites with single monitor all the time:\n")
unique(obs$poc[!cond]) 


pocs_on_sites <- obs[cond,] %>% 
    group_by(state_code, county_code, site_num, datum) %>%
    summarise(poc1 = sort(unique(poc))[1], 
              poc2 = sort(unique(poc))[2],
              poc3 = sort(unique(poc))[3],
              poc4 = sort(unique(poc))[4])
pocs_on_sites

sit %>% select(state_code, state_name) %>% unique %>%
    filter(state_code %in% unique(pocs_on_sites$state_code))

cat("Methods used at monitors with POC 9:\n")
met %>% filter(method_code %in% unique(obs$method_code[obs$poc==9])) %>%
    select(method_code, method_name, pollutant_standard)
```

Numbers from 1 through 4 present for sites with single monitor all time and 
are common for sites with many monitors for CO measuring.

Also, it was 4 monitors through all time at site number 41 present in table.
And 3 sites has 3 unique POC values, while it was just 2 sites with maximum 3
monitors at the same time. I guess, measurements was just transported from one
monitor to other.

I guess, sites have about fixed numbers of monitors with corresponding fixed
numbers. For example, I guess, there are 9 or more monitors in two sites at
California. California is advanced state, they can let such many monitors to
measure different values. And monitors with number 9 could both be chosen to
measure CO, probably due to good location or any other reasons. One monitor
with POC 9 was measuring with uncommon method
'INSTRUMENTAL - Gas Filter Correlation Teledyne API 300 EU'.
Probably such a monitor was chosen to experimentalize with method.

I guess presence of second site with CO monitor number 9 is randomness.
And **POC numbers 6, 7, and 8 aren't present because it wasn't a case**.

### Question 4

*Why there are more values with poc value 9 than for values 4 and 5?*

**POC 9 is more frequent than 5 because POC 5 is pretty randomly appeared and**
**at single site.**

POC 9 also was used in California to measure CO with as one of main monitors.
Monitor number 4 could be secondary at sites where it present.

### Question 5

*Whether event at some day is written into each observation at same day?*
